---
title: "Homework 4"
author: "Audrey Commerford"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Suppose *x*=1.1, *a*=2.2, and *b*=3.3. Assign each expression to the value of the variable z and print the value stored in z. 
```{r}
x <- 1.1
a <- 2.2
b <- 3.3

z <- x^(a^b)
print(z)

z <- (x^a)^b
print(z)

z <- 3*x^3 + 2*x^2 + 1
print(z)
```

2. Using the `rep` and `seq` functions, create the following vectors: 
```{r}
a <- c(1:8, 7:1)
print(a)

b <- rep(x=1:5, times=1:5)
print(b)

c <- rep(x=5:1, times=1:5)
print(c)
```

3. Create a vector of two random uniform numbers. In a spatial map, these can be interpreted as *x* and *y* coordinates that give the location of an individual (such as a marked forest tree in a plot that has been mapped). Using one of Rs inverse trigonometry functions (`asin()`, `acos()`, or `atan()`) convert these numbers into polar coordinates. 
```{r}
point <- runif(2)
print(point)
r <- (point[1]^2 + point[2]^2)^(1/2)
print(r)
theta <- atan(point[2]/point[1])
print(theta)
```

4. Create a vector `queue <- c("sheep", "fox", "owl", "ant")` where `queue` represents the animals that are lined up to enter Noah’s Ark, with the sheep at the front of the line. Using R expressions, update `queue` as:
the serpent arrives and gets in line;
the sheep enters the ark;
the donkey arrives and talks his way to the front of the line;
the serpent gets impatient and leaves;
the owl gets bored and leaves;
the aphid arrives and the ant invites him to cut in line.
Finally, determine the position of the aphid in the line.
```{r}
queue <- c("sheep", "fox", "owl", "ant")
queue <- c(queue, "serpent")
queue <- queue[-c(1)]
queue <- c("donkey", queue)
queue <- queue[-c(5)]
queue <- queue[-c(3)]
queue <- c(queue[1:2], "aphid", queue[3])
print(queue)
```
The aphid is third in line. 

5. Use R to create a vector of all of the integers from 1 to 100 that are not divisible by 2, 3, or 7.
```{r}
vector <- 1:100
vector <- which((vector %% 2) != 0 & (vector %% 3) != 0 & (vector %% 7 != 0))
print(vector)
```

