---
title: "Homework 5"
author: "Audrey Commerford"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Working with matrices, lists, and data frames

1. Assign to the variable `n_dims` a single random integer between 3 and 10.
Create a vector of consecutive integers from 1 to n_dims^2
```{r}
n_dims <- runif(n=1, min=3, max=10)
my_vec <- seq(from=1, to=n_dims^2)
head(my_vec)
```
Use the sample function to randomly reshuffle these values.
```{r}
new_vec <- sample(x=my_vec)
print(new_vec)
```
Create a square matrix with these elements.
```{r}
m <- matrix(data=new_vec, nrow=n_dims, ncol=n_dims)
```
Print out the matrix.
```{r}
print(m)
```
Find a function in r to transpose the matrix. Print it out again and note how it has changed.
```{r}
tm <- t(m)
print(tm)
```
Calculate the sum and the mean of the elements in the first row and then the last row.
```{r}
sum(tm[1,])
mean(tm[1,])
sum(tm[n_dims,])
mean(tm[n_dims,])
```
Read about the `eigen()` function and use it on your matrix
```{r}
eigen(tm)
```
Look carefully at the elements of values and vectors in the output. What kind of numbers are these? Dig in with the `typeof()` function to figure out their type.
```{r}
typeof(eigen(tm)$values)
typeof(eigen(tm)$vectors)
```
These are complex numbers. 

If have set your code up properly, you should be able to re-run it and create a matrix of different size because n_dims will change.

2. Create a list with the following named elements:

* `my_matrix`, which is a 4 x 4 matrix filled with random uniform values

* `my_logical` which is a 100-element vector of TRUE or FALSE values. Do this efficiently by setting up a vector of random values and then applying an inequality to it.

* `my_letters`, which is a 26-element vector of all the lower-case letters in random order.
```{r}
my_matrix <- matrix(data=runif(16), nrow=4, ncol=4)
my_logical <- runif(100)<0.5
my_letters <- sample(letters)
my_list <- list(my_matrix, my_logical, my_letters)
```
Then, complete the following steps:

* Create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.
```{r}
new_list <- list(my_matrix[2,2], my_logical[2], my_letters[2])
```
* Use the `typeof()` function to confirm the underlying data types of each component in this list
```{r}
typeof(new_list[[1]])
typeof(new_list[[2]])
typeof(new_list[[3]])
```
* Combine the underlying elements from the new list into a single atomic vector with the `c()` function. What is the data type of this vector?
```{r}
atomic_vector <- c(new_list[[1]], new_list[[2]], new_list[[3]])
print(atomic_vector)
typeof(atomic_vector)
```

3. Create a data frame with the two variables (= columns) and 26 cases (= rows) below:

* Call the first variable my_unis and fill it with 26 random uniform values from 0 to 10

* Call the second variable my_letters and fill it with 26 capital letters in random order.
```{r}
my_unis <- runif(n=26, min=0, max=10)
my_letters <- sample(LETTERS)
data_frame <- data.frame(my_unis, my_letters)
print(data_frame)
```
* For the first variable, use a single line of code in R to select 4 random rows and replace the numerical values in those rows with NA.
```{r}
data_frame[ ,1] <- my_unis
print(data_frame)

data_frame[sample(x=1:26, size=4, replace=TRUE),1] <- NA
print(data_frame)
```
* For the first variable, write a single line of R code to identify which rows have the missing values.
```{r}
which(!complete.cases(data_frame))
```
* Re-order the entire data frame to arrange the second variable in alphabetical order
```{r}
data_frame <- data_frame[order(data_frame$my_letters),]
print(data_frame)
```
* Calculate the column mean for the first variable.
```{r}
mean(data_frame$my_unis, na.rm=TRUE)
```
